\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{comment}
\usepackage{todonotes}
\usepackage{msc5}
\usepackage{listings}
\usepackage{url}
\usepackage{tikz-qtree}
\usepackage{graphicx}
\usepackage{paralist}
\usepackage{algorithm,algorithmic}
\usepackage{amsmath, amssymb, amsthm, amsfonts, caption, stmaryrd, mathtools, syntax, mdframed}
\usepackage{float}
\usepackage{cryptocode}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\title{QUIC Handshake Key Discard Alternatives}
\author{Eric Rescorla}
\date{\today}

\begin{document}

\maketitle

\section{Status Quo}
\label{sec:status-quo}


\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [DATA]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [DATA, ACK]}}{sr}{cl}
  \nextlevel
  \action*{Handshake Confirmed; Drop Handshake Keys}{cl}
  \nextlevel[3]  
  \mess{\emph{1-RTT [ACK]}}{cl}{sr}
  \nextlevel    
  \action*{Handshake Confirmed; Drop Handshake Keys}{sr}
  \nextlevel[2]
  
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Basic QUIC handshake (without 0.5 RTT data).}
\label{fig:basic-handshake}
\end{figure}

Figure~\ref{fig:basic-handshake} shows the basic QUIC handshake along
with the various checkpoints. This just for reference and hopefully
non-controversial.

This design is known to have several problems, specifically:

\begin{enumerate}
\item A dropped ACK can cause the client to retransmit CFIN
  indefinitely.
\item It is possible for the server to have handshake data
  outstanding when the client initiates migration. 
\end{enumerate}

We show these scenarios below.

\subsection{Indefinite CFIN Retranmission}
\label{sec:indef-cfin-retr}

As shown in Figure~\ref{fig:cfin-retransmit}, if the server's ACK of CFIN
gets lost but the server sends 1-RTT data and has it ACKed, then the
server will drop the handshake keys and the client will indefinitely
retransmit CFIN, at least until it sends 1-RTT data and has it ACKed.
[TODO: I believe that with the changes to recovery, if the client
sends 1-RTT data it will be transmitted, but I need to check. Ian?]


\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \dummyinst{drop}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [ACK]}}{sr}{drop}
  \nextlevel[2]  
  \mess{\emph{1-RTT [DATA]}}{sr}{cl}
  \nextlevel[2]
  \mess{\emph{1-RTT [ACK]}}{cl}{sr}
  \nextlevel    
  \action*{Handshake Confirmed; Drop Handshake Keys}{sr}
  \nextlevel[2]
  \action*{IGNORED}{sr}
  \nextlevel[.5]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{CFIN Retransmission with dropped ACK}
\label{fig:cfin-retransmit}
\end{figure}


\subsection{Outstanding Handshake Data During Migration}
\label{sec:outst-handsh-data}

Consider the situation shown in 
\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [DATA]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [ACK]}}{sr}{cl}
  \nextlevel
  \action*{Handshake Confirmed; Drop Handshake Keys}{cl}
  \nextlevel[2]  
  \mess{\emph{MIGRATION, 1-RTT [DATA]}}{cl}{sr}  
  \nextlevel  
  \action*{Handshake data still outstanding!}{sr}
  
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Outstanding handshake data during migration}
\label{fig:outstanding-handshake-migration}
\end{figure}

The issue here is that the server's signal that the handshake
is done is an ACK of its 1-RTT data, but there's no guarantee
that this precedes the client's migration data, because the
client can migrate as soon as it has received an ACK of its
1-RTT data. Even if the server sends its own 1-RTT data in
the same packet as its ACK, there is no requirement that
the client ACK at the same time as it sends its migration
data.

Note that the converse case cannot happen: the client cannot
initiate migration until handshake confirmed and it is required
to drop the handshake keys at that point.

\textbf{QUESTION}: Do current implementations (especially clients)
ACK handshake packets prior to handshake confirmed? If so,
this case ought to be quite rare.


\section{Never Drop the Keys (PR 3121)}
\label{sec:never-drop-keys}

The basic idea here is to (1) never drop the Handshake keys
and (2) continue to send ACKs for packets you have received
(as agreed upon separately in YUL, you MUST ACK each
packet at least once). The result is that the handshake
goes quiescent even if no application data is ever transmitted,
though you will not reach ``handshake confirmed'' in this
case, as shown in Figure~\ref{fig:never-drop-1}.

\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN, ACK]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[3]
  \mess{\emph{Handshake [ACK]}}{sr}{cl}
  
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Never drop keys (no data sent)}
\label{fig:never-drop-1}
\end{figure}


One obvious question is what happens in the case where you
\emph{are} sending data. In the basic case, we would just
expect ACK piggybacking on the data, as shown in Figure~\ref{fig:never-drop-2}

\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN, ACK]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [DATA]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [ACK], 1-RTT [DATA, ACK]}}{sr}{cl}
  \nextlevel
  \action*{Handshake Confirmed}{cl}
  \nextlevel[3]  
  \mess{\emph{1-RTT [ACK]}}{cl}{sr}
  \nextlevel    
  \action*{Handshake Confirmed}{sr}
  \nextlevel[2]
  
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Never drop keys (data sent)}
\label{fig:never-drop-2}
\end{figure}

You can also safely implement implicit ACK. At the point
where you have handshake confirmed, you can stop sending
any of your Handshake DATA frames (though you still
need to ACK any of the peer's Handshake frames, or you
will get deadlocks if the peer doesn't implement implicit
ACKs.)

Kazuho had raised two concerns about this design:

\begin{enumerate}
\item Because handshake confirmed is not guaranteed, key update
  is confusing.
\item It is possible to have outstanding handshake data at the
  point where you want to do migration.
\end{enumerate}

I don't really understand point (1). As noted in
\url{https://github.com/quicwg/base-drafts/issues/3212}, the handshake
confirmed test for key update is redundant, and so we can just re-code
this as ``don't initiate key update for key $n+1$ until you are sure
the other side has key $n$''. And this just lets us have the usual
rule without a special case for the handshake $\rightarrow$ 1-RTT
transition.

Point (2) is the same issue as is raised in
Section~\ref{sec:outst-handsh-data} above. It is possible that this
case is less likely with the ``never drop keys'' approach because
the client is required to send ACKs for handshake data. Thus, if
the client piggybacks ACKs for the server's handshake packets
on its own CFIN (which is natural), the server will never be
able to process 1-RTT packets from the client when handshake
data is outstanding, in which case the problem goes away. Note
that this is separate from the packet loss edge cases, because
the client can guarantee this condition. Note that the server
can still receive late copies of the CFIN, and would
need to be required to ACK them on the original path.


\section{Server sends HANDSHAKE_DONE(PR 3145)
}
\label{sec:serv-sends-handsh}

PR 3145 suggests an explicit signal from the server that the handshake
is done. This signal, unlike the ACK for the CFIN, is reliably
retransmitted. Figure~\ref{fig:server-hd} shows this.


\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [HANDSHAKE\_DONE]}}{sr}{cl}
  \nextlevel
  \action*{Handshake Confirmed; Drop Handshake Keys}{sr}
  \nextlevel[2]
  \action*{Handshake Confirmed; Drop Handshake Keys}{cl}
  \nextlevel[3]  
  \mess{\emph{1-RTT [ACK]}}{cl}{sr}
  \nextlevel    
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Server sends HANDSHAKE\_DONE}
\label{fig:server-hd}
\end{figure}

This addresses the migration issue because the server (or at least
a compliant server) will not send HANDSHAKE\_DONE before it has
received CFIN. Thus, the client will not consider the handshake
confirmed if the server has any outstanding handshake data,
Similarly, the client cannot send any handshake data after handshake
confirmed.

Note that this diagram shows that the handshake is confirmed on the
server 1 RTT faster than with either of the two previous cases,
but equivalent performance can be achieved by having both
sides send 1-RTT PINGs as soon as they have 1-RTT keys,
at which point the handshakes will be confirmed at the same
time (at receipt of the CFIN on the server, and 1-RTT later
on the client).

[TODO: Make diagram]


\section{Server drops keys on receipt of 1-RTT Packet}
\label{sec:server-drops-keys}

In Ian's doc. he suggested that we require the server to drop
handshake keys when it receives a 1-RTT packet, i.e., one RTT
sooner than the current design.

\begin{figure}[H]
\begin{center}
\resizebox{0.8\columnwidth}{!}{\begin{msc}{}
  \drawframe{none}
  \setlength{\topheaddist}{0cm}
  \setlength{\instdist}{8.5cm}
  \setlength{\instwidth}{1.75cm}
  \declinst{cl}{}{Client $C$}
  \declinst{sr}{}{Server $S$}

  \mess{\emph{Initial [C]}}{cl}{sr}
  \nextlevel[2]
  \mess{\emph{Initial [SH], Handshake[EE...SFIN]}}{sr}{cl}
  \nextlevel
  \action*{Server knows 1-RTT keys}{sr}
  \nextlevel[2]
  \mess{\emph{Handshake [CFIN]}}{cl}{sr}
  \nextlevel
  \action*{Client knows 1-RTT keys; Handshake Complete}{cl}
  \action*{Handshake Complete}{sr}
  \nextlevel[2]
  \mess{\emph{1-RTT [DATA]}}{cl}{sr}
  \nextlevel[2]  
  \action*{Handshake Confirmed; Drop Handshake Keys}{sr}  
  \nextlevel[2]
  \mess{\emph{1-RTT [ACK]}}{sr}{cl}
  \nextlevel
  \action*{Handshake Confirmed; Drop Handshake Keys}{cl}
  \nextlevel[2]
\end{msc}
}
\end{center}
\caption{Server drop on 1-RTT}
\label{fig:server-drop-on-10rtt}
\end{figure}


This solves the migration problem because the server will no
longer send handshake packets (or ACKs) to the client after
receiving the first 1-RTT packet, and the client is forbidden
to initiate migration prior to having sent and received a 1-RTT
packet. However, it recreates the problem with the status
quo which is that if the client never sends 1-RTT data but
the server does, you can end up with the client indefinitely
retransmitting ACKS.







\end{document}

